### 2009-12-10 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Revisions of chapter 5 from Niki... This should have been committed on the last revision.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@690 559edde2-a4d4-44e0-96fc-4035e40d7396

Providing data persistence for an existing sensor networks require analyzes of
the infrastructure and the nature of the produced data. Based on the taxonomies
of chapter 3, and the requirements of a data persistence for NetBEAMS in
chapter 4, this chapter aims at analyzing a group of database technologies
selected from literature review topics covered in chapter 2. Similarly,
supported by the scope of the required implementation presented in this
chapter, the comparison of the technogies is summarized, highlighting the
selected database system used for the design of a component for NetBEAMS and
used for the design of the experiments.

\section{Scope of the use of the Techonolgy}

The scope of the use of the database can involve different functional and
non-functional requirements defined in the last chapter. This work does not
intend to implement an application based on the database system selected, but
to provide the foundation of a database management system that can be used to
develop one focused in less refactoring and high scalability and performance.
For this reason, the use of the database system will be restricted to the data
persistence and use through native system or through the use of native
programming languages or API. In this way, the scope of this technology
selection can be summarized as follows:

\begin{itemize}
  \item Provide data management over the collected sensor data, as a
  conventional database system;
  \item Must be able to scale in the face of an increase to the data load of
  the sensor network, as described in chapter 2;
  \item  Must be able to use single or multiple servers to improve performance
  and optionally deal with Data-Centric approaches to sensor data partitioning;
  \item Must provide a data model that scales, decreasing the constant data
  schema changes.
\end{itemize}

@@

One of the essential questions related to data persistence is regarding the
data model and breed of the database. That is, the use of the relational or
any other model to describe data along with its management. Traditionally, the
relational data model has been used to provide data persistence for different
types of projects, inclusing sensor networks \cite{sn-dataware-house,
db-xml-enabled, sn-db-tinydb}. However, tradition does not always get translated
to efficiency in terms of problem solving. For this reason,
\cite{db-is-rdbs-dommed} reflects about the use of alternatives to the
relational model when the data persistence problem in consideration is to
provide data persistence for dynamic environments as Internet applications. The
author also suggests that the relational model is hard to maintain in
ever-changing environments, and thus, proposes the use of the Key-Value-Pair
(KVP) data model and a different number of technologies. Similarly,
\cite{cloud-comp-survey} reports similar trends in academic and industrial
research in distributed computing have used KVP databases in powerful systems
such as Cloud Computing \cite{cloud-comp-architectures}. For this reason, in
addition to the different database systems used by projects in the
literature review, this work also includes the review of a KVP database system
called mongoDB \cite{mongodb}, since it was among those cited in the surveys.
In such a manner, the list of the systems considered for assessment is the
following:

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  dateware house with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks
  implementation such as \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source KVP database
  system suggested by different surveys to be alternatives to relational
  databases \cite{db-is-rdbs-dommed, cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. Furthermore, it was selected because
  NetBEAMS represents serialized objects in XML, as shown in section
  \ref{sec:dsp-message}.
\end{itemize}

The following sections reviews these technologies' capabilities to adhere to
the specifications of the taxonomies defined in chapter 3, as well as the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the data sink
  at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for
  distribution using the OPEnDAP format;
  \item An automated approach to the data collection procedures is performed 
  using NetBEAMS \cite{netbeams2009}, using software components to obtain and
  send the data to the data sink node. A complete description of this approach
  is detailed in section \ref{sec:netbeams-architecture}.
\end{itemize}

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}. After the
collected data captured from the sensors, it requires further processing
in order to be archived for later reuse. Different types of users are expected
to use the collected data, ranging from researchers, students, and the public 
over the Internet.

@@

Since all of developed database systems were implemented with the intension of
persisting data in long-term storage devices, every single one in the list
provides support for data persistence and management. Therefore, they are
compliant to the purpose of Data Archival for the sensor networks.

@@

As shown in the previous chapter, NetBEAMS's architecture is based on sensor
network whose topology is a single-hop star with a centralized data sink. In
this way, all the collected data is transmitted to that single location
\ref{sec:sn-infrastructure}. In the same vein, the simplest location to save
the collected data is through the use of at least an \textbf{External Storage}.
Similarly, the use of a Data-Centric Storage strategy is considered in
situations where the data sink is about to reach its storage capacity. As a
consequence, the \textbf{Data-Centric Storage} separates the collected data
into dissimilar locations, based on any property of the data. Therefore,
there are two different approaches to data partitioningThe two different
flavors:

@@

The External Storage approach is the most common way to implement persistence
for any breed of systems since data management is usually offered by most
database systems out-of-the-box.

@@

The most common problem related to a single External Data Storage is that 
resources may run into data management problems such reaching the disk storage
capacity or hardware failure.

@@

The benefit of the Data-Centric approach can be directly associated with
database partitioning mechanisms described earlier. Since the collected data is
distributed into different locations, this approach decreases the size of
datasets when the database server performs a query processing in a database
table. For this reason, this approach helps the database administrators to
scale their database infrastructure and improve the database performance.

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-partitioning-relational}. Both the former and the
latter are used in regular schema-dependent models or schema-less models,
which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas, db-partitioning-relational}.

\begin{itemize}
  \item \textbf{Directory Based Partitioning}: an ordinary lookup service is
  used to identify which database shard to use, where the lookup table can be
  stored in a secondary area;
  \item \textbf{Key or Hash Based Partitioning}: the value of a given key
  property can  be used as the key of the shard. Moreover, it can be used as 
  an input to a hash function, whose output determines at which database shard
  the data needs to be stored.
\end{itemize}

An example of a Data-Centric storage with different shards can be see in Figure
\ref{fig:database-sharding-by-region}, where the collected data is segregated
by its origin. Each shard is assumed to be placed in different server
hosts, as well as each of them uses the same ``denormalized'' data models.

@@

\begin{itemize}
  \item \textbf{Rebalancing Shards}: considering a schema-dependent database
  system, changes on the schema of one shard requires rebalancing the
  collected data on each of the shards. As a consequence, the database shards
  transfers data to new locations whenever necessary. This technique is
  starting to appear as COTS implementations such as MySQL \cite{mysql} and
  mongoDB \cite{mongodb};
  \item \textbf{Referential integrity}: also considering a schema-dependent
  database system, any cross-shard query may fall into the problem where
  references do no exist in other shards, since the application layer is
  responsible for enforcing data integrity. In this way, examples of such
  include verifying constraints of foreign keys when the partition is done
  using Vertical table partitioning, considering that data collections are
  organized in separate spaces, and physically placed in different shard nodes
  \cite{db-partitioning-relational}.
\end{itemize}

@@

All the selected database systems support the External Storage method. However,
they all differ with the support to Data-Centric Storage support. Therefore,
MySQL and mongoDB are promising choices, with the difference of the use of the
data model.

@@

One of the most common practices in the area of database system is to use the
relational model to persist data, although the application of the system may
not fit to solve the problem. In fact, the main users of sensor networks may
not possess any expertise in database systems or data modeling, based on the
fact that they are from different science areas. \cite{sn-programming-language}
suggested the use of programming languages abstraction instead of regular
database systems for researchers. For instance, SF-BEAMS is managed by Marine
Biologists without expertise in Data Management, Modeling systems, and for
this reason, one of the requirements of the system is trying the use of
schema-less approaches. The \textbf{Key-Value-Pair Data Model} or the
\textbf{Document-Oriented Data Model} seems to be the simplest choices of the
models.

@@

Considering the inception of a relational data model \cite{relational-model}
and the use of the YSI Sonde data as the main entity in the system, let Figure
\ref{fig:Relational-Model-Original} represent a prototype of the relational 
model, after passing through the process of normalization
\cite{db-normalization}.

@@

Supposing a new type is added into the system, let the refactored version of
the relational model be depicted in Figure
\ref{fig:Relational-Model-Addition-Modified}. Some observations are
considered in this scenario:

@@

\begin{itemize}
  \item Data schema that constantly changes require constant database
  normalization processes, changes to structure, database management, and all
  the code that uses the data. Schema-Dependent databases always require
  refactoring of the schema to accommodate changes to the entites' properties;
  \item The approach of Relational Model might not fit the nature of the
  collected data from sensor devices, since they tend to change over time. 
  Furthermore, the support to time-series or provenance are not better
  supported \cite{sn-provenance};
  \item Some research have suggested changes to the SQL clauses to better
  support time-series when collecting data using the relational model 
  \cite{sn-db-newop}.
\end{itemize}

@@

Although this data model has been used in different types of applications,
developers have tried to use the Relational Model to design an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. As mentioned earlier, with the
emergence of Internet applications, the necessity of such a model that does
not require constant refactoring led developers to propose such manner using
the relational model as documented in technical articles such as
\cite{db-kvp-in-relational01} and \cite{db-kvp-in-relational02}. Based on
these articles, consider the relational model depicted in Figure
\ref{fig:KVP-on-Relational-Model} as a data model for persisting data to our
case study using the key-value strategy. It is clear that this approach
contains problems: repetition occurs, since the nature of
time-series data requires the use of a timestamp key for each property of the sensor. Therefore, this strategy is not
well-suited to provide persistence data from collected sensor networks.

@@

There are no records of the use of this data model with Sensor Networks.
Different variants of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure, freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json}. mongoDB implements the
following:

@@

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describes the data, depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and is compared on table
\ref{tab:schema-vs-schemaless}:

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model is defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships \cite{db-normalization};
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:schema-vs-schemaless-daccess}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic operations CRUD\footnote{Create-Retrieve-Update-Update
      database operations} data are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, contains functions for aggregation and complex filter.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via API\footnote{Application
      Programming Interface} through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like =, !=, <, > that ca be applied;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:tab:schema-vs-schemaless-api-interface}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC\footnote{Open Database
      Connectivity};
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Systems tend to provide SOAP/REST \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

Since the selected data model is the schema-less one, the following can be seen
from the technologies:

\begin{itemize}
  \item \textbf{MySQL}: supports the Relational Data Model;
  \item \textbf{TinyDB}: supports the Relational Data Model; 
  \item \textbf{MongoDB}: supports Key-Value Pair Data Model;
  \item \textbf{IBM DB2}: supports the Relational Model and the Structured
  Data models.
\end{itemize}

@@

\textbf{Data Description} is the methods used to describe the collected data
from sensors, specially those produced by real-time data streams. Since those
type of data does not include descriptions of its values, data provenance
offers guidelines to improve the description of the collected data.

An additiona to data provenance, annotations is another technique used to
better describe the collected data from sensor devices. Its main purposes is to
provide means of identifying observations, creating indexes based on simple
keywords. For an environmental sensor networks, this is a valuable type of value
used to better manage the collected data.

@@

The injection of provenance data into the collected data enriches description
of a given event, and is always useful to determine different aspects of the
data. As mentioned in chapter 3, data provenance helps identifying the nature
of the data by the identification of the properties of a given sensor device,
for example. Similarly, provenance adds spatio-time data in order to correlate
the observed data with time a given location and time of occurrence.

@@

Additional data being inserted together with the raw data collected from the
sensor devices. This additional data may contribute with any decreasing number
of disk storage for the variable defined. Therefore, the decreasing amount
of disk space may be one the reasons to not use annotations.

@@

This type of data does not make part of the collected data, nor provenance
information.

@@

As a direct result from the previous section, the use of \textbf{Centralized Query
Processing} is indicated for NetBEAMS, also given the fact that SF-BEAMS data
goes directly to the individual network sink and, therefore, a centralized
point of the data.

@@

When the Centralized Query processing is used, problems may occur. Depending on
the size of the datasets and the database technology, performance may be a
concern. For example, this procedure creates of the alleged Funneling Affect
\cite{sn-storage04}, since the point-of-traffic is concentrated in the
centralized system.

@@

All the database systems technology support a centralized query system. In
addition to that, they can be distributed into different distributed nodes when
used in a partioned environement.

@@

Sensor Network data can be saved in either Centralized or Distributed Systems.
While Centralized Database Systems are easier to manage, it may
face challenges regarding its data. In order to implement a Data-Centric
solution, a distribute database system must be in place in order to manage
the different data by a single node.

\subsection{Advantages of Database Partitioning}

\begin{itemize}
  \item Data-Centric queries and data use;
  \item Solves bottleneck problems related to reads/writes;
  \item Decrease the funneling effect by directing queries to given data
  partition;
\end{itemize}

\subsection{Disadvantages of Database Partitioning}

\begin{itemize}
  \item Very advanced topics in Database System;
  \item Partitions need to be rebalanced in case the key of the selected 
  partitioning strategy is changed.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports Data Replication;
  \item \textbf{TinyDB}: NO Supports to distributed data;
  \item \textbf{MongoDB}: Supports Data Replication;
  \item \textbf{IBM DB2}: Supports Data Replication.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Is  open-source, supports hot backup and contains
  scripting APIs from  the community;
  from the community. Also APIS are available for the majority of languages;
  \item \textbf{TinyDB}: Not open-source;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from  the community, offering native APIS in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not Open-Source, but offers support from its
  community.
\end{itemize}

Only MySQL and MongoDB supports the non-functional requirements defined in this
section.

>>>>

Providing data persistence for an existing sensor network requires analysis of
the infrastructure and the nature of the data produced. Based on the
taxonomies of Chapter 3, and the requirements of a data persistence for
NetBEAMS in Chapter 4, this chapter aims at analyzing a group of database
technologies selected from the topics covered in Chapter 2.
Similarly, supported by the scope of the required implementation presented in
this chapter, the comparison of the technologies is summarized, highlighting
the selected database system used for the design of a component for NetBEAMS
and used for the design of the experiments.

\section{Scope of the use of the Technology}

The scope of the use of the database can involve different functional and
non-functional requirements defined in the last chapter. This work does not
intend to implement an application based on the database system selected, but
to provide the foundation of a database management system that can be used to
develop one focused less in refactoring and high scalability and performance.
For this reason, the use of the database system will be restricted to the data
persistence and use through a native system or through the use of native
programming languages or API. In this way, the scope of this technology
selection can be summarized as follows:

\begin{itemize}
  \item Provide data management over the collected sensor data, as a
  conventional database systems;
  \item Must be able to scale in any increases to the data load of
  the sensor network, as described in chapter 2;
  \item  Must be able to use single or multiple servers to improve performance
  and optionally deal with Data-Centric approaches to sensor data partitioning;
  \item Must provide a data model that scales, decreasing the constant data
  schema changes.
\end{itemize}

@@

One of the essential questions related to data persistence is regarding the
data model and types of the database. That is, the use of the relational or
any other model to describe data along with its management. Traditionally, the
relational data model has been used to provide data persistence for different
types of projects, including sensor networks \cite{sn-dataware-house,
db-xml-enabled, sn-db-tinydb}. However, tradition does not always get
translated into efficiency in terms of problem solving. For this reason,
\cite{db-is-rdbs-dommed} reflects about the use of alternatives to the
relational model when the data persistence problem in consideration is to
provide data persistence for dynamic environments such as Internet applications.
The author also suggests that the relational model is hard to maintain in
ever-changing environments, and thus, proposes the use of the Key-Value-Pair
(KVP) data model and a different number of technologies. Similarly,
\cite{cloud-comp-survey} reports similar trends in academic and industrial
research in distributed computing using KVP databases in powerful systems
such as Cloud Computing \cite{cloud-comp-architectures}. For this reason, in
addition to the different database systems used by projects in the
literature review, this work also includes the review of a KVP database system
called mongoDB \cite{mongodb}, since it was among those cited in the surveys.
In such a manner, the list of the systems considered for assessment is the
following:

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  data warehouse with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in Section 
  \ref{sec:sn-persitence-storage}, TinyDB is a database system used in many
  different sensor networks implementation such as \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source KVP database
  system suggested by different surveys to be alternatives to relational
  databases \cite{db-is-rdbs-dommed, cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. Furthermore, it was selected because
  NetBEAMS represents serialized objects in XML, as shown
  \cite{netbeams-dsp-architecture}.
\end{itemize}

The following sections review these technologies' capabilities to adhere to
the specifications of the taxonomies defined in chapter 3, as well as the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI Sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the data sink
  at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for
  distribution using the OPeNDAP format;
  \item An automated approach to the data collection procedures is performed 
  using NetBEAMS \cite{netbeams2009}, using software components to obtain and
  send the data to the data sink node. A complete description of this approach
  is \cite{netbeams-dsp-architecture}.
\end{itemize}

The nature of data from NetBEAMS through SF-BEAMS is used for the purpose of
\textbf{Data Archival}. After the collected data captured from the sensors, it
requires further processing in order to be archived for later reuse. Different
types of users are expected to use the collected data, ranging from
researchers, students, and the public over the Internet.

@@

Since all of developed database systems were implemented with the intention of
persisting data in long-term storage devices, every one listed provides
support for data persistence and management. Therefore, they are compliant to
the purpose of Data Archival for the sensor networks.

@@

As shown in the previous chapter, NetBEAMS's architecture is based on sensor
network whose topology is a single-hop star with a centralized data sink. In
this way, all the collected data is transmitted to that single location
\ref{sec:sn-infrastructure}. The simplest location to save the collected data
is through the use of at least an \textbf{External Storage}. Similarly, the
use of a Data-Centric Storage strategy is considered in situations where the
data sink is about to reach its storage capacity. The \textbf{Data-Centric
Storage} separates the collected data into dissimilar locations, based on any
property of the data. Therefore, there are two different approaches to data
partitioning:

@@

The External Storage approach is the most common way to implement persistence
for any type of systems since data management is usually offered by most
database systems out-of-the-box.

@@

The most common problems related to a single External Data Storage approach are
related to the capacity of the device and operational problems. If the
external storage device is situated in the network, then the capacity of the
device may be limited to receive collected data from the sensors in the
network, since it may run out of space. Similarly, storage devices are very
error prone and its hardware part may fail.

@@

The benefit of the Data-Centric approach can be directly associated with
database partitioning mechanisms described earlier. Since the collected data is
distributed into different locations, this approach decreases the size of
datasets when the database server performs a query processing in a database
table. For this reason, this approach helps the database administrators scale
their database infrastructure and improve the database performance.

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-partitioning-relational}. Both the former and the latter are used in
regular schema-dependent models or schema-less models. Different approaches
worth mentioning:

\begin{itemize}
  \item \textbf{Hash Partitioning}: a hash key is computed based on
  different table columns of an entity \cite{db-shard-schemas,
  db-partitioning-relational}, providing an even distribution of the data;
  \item \textbf{Key Partitioning}: is based on the hash partitioning over the
  values of a specific key \cite{db-mongo-partition};
  \item \textbf{Range Partitioning}: the partition is based on specific ranges
  of the values of given a given column of a table. For instance, the collected
  data from the sensors could be separated by the year of its collection. For
  example, data collected in the year of 2009 would be stored in a different
  partition than the data collected in the year 2010
  \cite{db-partitioning-relational};
  \item \textbf{Directory Based Partitioning}: an ordinary lookup service is
  used to identify which database partition to use \cite{db-shard-schemas}.
\end{itemize}

An example of a data-centric storage with different partitions can be seen in
Figure \ref{fig:database-sharding-by-region}, where the collected data is
partitioned by its origin. Each shard is assumed to be placed in different
server hosts, as well as each of them uses the same ``denormalized'' data
models.

@@

\begin{itemize}
  \item \textbf{Rebalancing Partitions}: considering a schema-dependent database
  system, changes on the schema of one shard requires rebalancing the
  collected data on each of the shards. As a consequence, the database shards
  transfers data to new locations whenever necessary. This technique is
  starting to appear as COTS implementations such as MySQL \cite{mysql} and
  mongoDB \cite{mongodb};
  \item \textbf{Referential integrity}: also considering a schema-dependent
  database system, any cross-shard query may run into the problem where
  references do no exist in other shards, since the application layer is
  responsible for enforcing data integrity. In this way, examples of such
  include verifying constraints of foreign keys when the partition is done
  using Vertical table partitioning, considering that data collections are
  organized in separate spaces, and physically placed in different shard nodes
  \cite{db-partitioning-relational}.
\end{itemize}

@@

All the selected database systems support the External Storage method. However,
they all differ somewhat in terms of Data-Centric Storage support. Therefore,
MySQL and mongoDB are promising choices, with the difference of the use of the data model.

@@

One of the most common practices in the area of database system is to use the
relational model for data persistence, although the application of the system
may not fit to solve the problem. In fact, the main users of sensor networks may
not possess any expertise in database systems or data modeling, based on the
fact that they are from different science areas. \cite{sn-programming-language}
suggested the use of programming languages abstraction, instead of regular
database systems, to provide access to the collected data for both
technical and non-technical users. For instance, marine biologists manage
SF-BEAMS without expertise in data management or modeling systems. For this
reason, one of the requirements of the system is the use of schema-less
approach, which describes the collected data without data modeling process
used in schema-dependent models. The \textbf{Key-Value-Pair Data Model} seem
to be an optional candidate that provides such a method of data design.

@@

Considering the inception of a relational data model \cite{relational-model}
and the use of the YSI Sonde data as the main entity in the system, Figure
\ref{fig:Relational-Model-Original} represents a prototype of the relational 
model, after passing through the process of normalization
\cite{db-normalization}.

@@

When a new type is added into the system, the refactored version of the
relational model be depicted in Figure
\ref{fig:Relational-Model-Addition-Modified}. Some observations are considered
in this scenario:

@@

\begin{itemize}
  \item Data schema that constantly change requires continuous database
  normalization processes, changes to structure, database management, and all
  the code that uses the data. Schema-Dependent databases always require
  refactoring of the schema to accommodate changes to the entities' properties;
  \item \cite{sn-provenance} describes the relational model as problematic 
  when describing collected data from sensor devices;
  \item \cite{sn-db-newop} suggests changes to the SQL clauses of TinyDB to
  better support time-series, an important property when describing.
\end{itemize}

@@

Although this data model has been used in different types of applications,
developers have tried to use the Relational Model to design an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. As mentioned earlier, with the
emergence of Internet applications, the necessity of such a model that does
not require constant refactoring led developers to propose the relational model
as documented in technical articles such as \cite{db-kvp-in-relational01} and
\cite{db-kvp-in-relational02}. Based on these articles, consider the
relational model depicted in Figure \ref{fig:KVP-on-Relational-Model} as a
data model for persisting data to our case study using the key-value
pair strategy. It is clear that this approach contains problems: repetition
occurs, since the nature of time-series data requires the use of a timestamp
key for each property of the sensor. Therefore, this strategy is not well
suited to provide persistence data from collected sensor networks.

@@

There are no records of the use of this data model with Sensor Networks.
A different variant of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure, freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json}. mongoDB implements the
following:

@@

For instance, all data needed for the YSI Sonde Data, as well as all necessary
provenance data that describes the data depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and is compared on Table
\ref{tab:schema-vs-schemaless}:

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model is defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships \cite{db-normalization};
      \item Data Provenance can be provided through data types such as
      ``timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enables them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:schema-vs-schemaless-daccess}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic CRUD (Create-Retrieve-Update-Delete)
      database operations are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, containing aggregation functions to apply on the returned values.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via an API (Application
      Programming Interface) through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like equals to (=), different than (!=), less than
      (<), greater than (>), among others;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:tab:schema-vs-schemaless-api-interface}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC (Open Database
      Connectivity);
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Tend to provide Web Services Interfaces such as the REST
      \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

The following summarizes which data model each of the candidate database
systems use.

\begin{itemize}
  \item \textbf{MySQL}: uses the Relational Data Model with the use of SQL;
  \item \textbf{TinyDB}: uses the Relational Data Mode with the use of SQL; 
  \item \textbf{MongoDB}: uses Key-Value Pair Data Model with an Non-Structure
  Query language in forms of programming language abstraction;
  \item \textbf{IBM DB2}: uses the Relational and Structured models with SQL
  and XPath, respectively.
\end{itemize}

@@

\textbf{Data Description} is the method used to describe the collected data
from sensors, especially those produced by real-time data streams. Since that
type of data does not include descriptions of its values, data provenance
offers guidelines to improve the description of the collected data.

In addition to data provenance, annotation is another technique used to
better describe the collected data from sensor devices. Its main purpose is to
provide a means of identifying observations, creating indexes based on simple
keywords. For an environmental sensor network, this is invaluable used to
better manage the collected data.

@@

The inclusion of provenance data into the collected data enriches the
description of a given event, and is always useful to determine different
aspects of the data. As mentioned in chapter 3, data provenance helps identify
the nature of the data by the identification of the properties of a given
sensor device, for example. Similarly, provenance adds spatial-time data in
order to correlate the observed data with time a given location and time of
occurrence.

@@

Additional data being inserted together with the raw data collected from the
sensor devices may contribute with a decreasing amount of disk storage. 
Therefore, the use of provenance-away data increases the required disk space
used to save the collected data from sensor devices.

@@

This type of data does not make up part of the collected data, nor provenance
information.

@@

As a direct result from the previous section, the use of \textbf{Centralized Query
Processing} is indicated for NetBEAMS, because the SF-BEAMS data goes directly
to the individual network sink and, therefore, a centralized point of the data.

@@

When the Centralized Query processing is used, problems may occur. Depending on
the size of the datasets and the database technology, performance may be a
concern due to the alleged Funneling Affect \cite{sn-storage04}, since the
point-of-traffic is concentrated in the centralized system.

@@

All the database systems technology support a centralized query system. In
addition to that, they can be sorted into different distributed nodes when
used in a partitioned environment.

@@

Sensor Network data can be saved into database system that are organized
by either a Centralized or Distributed System. While a centralized database
system may be considered to be easier to manage, it may fail to adjust to
changes of requirements to manage data. On the other hand, a distribute
database system can be used to implement a data-centric solution. Therefore, a
database system that can be organized in both approaches is a good candidate to
be used a data persistence technology for sensor networks.

\subsection{Advantages of Distributed Systems}

Distributed Systems can improve the performance of data management by providing
techniques for parallel data processing such as Map Reduce \cite{map-reduce,
bigtable}. It also helps decreasing the system load and bottleneck problems
related data creation and retrieval. It is also seen as an optional mechanism
of reducing the funneling affect of a centralized system by directing queries
to a given data partition \cite{sn-storage04}. Data Replication
\cite{data-replication} provides a ways to decentralize data access by
providing duplicated data that is intended to be used on reading operations.

\subsection{Disadvantages of Distributed Systems}

Requires more technical knowledge expertise to manage the different systems
that composed a distributed database.

@@

\begin{itemize}
  \item \textbf{MySQL}: provides support to data replication;
  \item \textbf{TinyDB}: it does not provide support to data replication;
  \item \textbf{MongoDB}: supports data replication;
  \item \textbf{IBM DB2}: supports data replication.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Is  open-source, supports hot backup and contains
  scripting APIs from  the community;
  from the community. Also APIs are available for the majority of languages;
  \item \textbf{TinyDB}: Not open-source;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from  the community, offering native APIs in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not Open-Source, but offers support from its
  community.
\end{itemize}

Only MySQL and mongoDB supports the non-functional requirements defined in this
section.
